---
title: "A generalized format for longform travel data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Travel_data_template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(mobility)
library(ggplot2)
library(reshape2)
library(viridis)
library(ggstance)
library(knitr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  autodep=TRUE
)
```

There are many sources of travel data that researchers wish to fit models to. So, we have designed a generalized data frame template to standardize travel data from various sources into a longform format that is compatible with the modeling and simulation tools in this package. The \code{travel_data_sim} object contains simulated content to illustrate the structure of the data. This example data set contains simulated values of location information and observed number of trips *among* origin and destination locations and *within* home locations.
```{r}
str(travel_data_sim)
```

```{r, echo=FALSE}
tab <- matrix(
  c('date_start', 'date', 'beginning of the time interval for the trip count',
    'date_stop', 'date', 'end of the time interval for the trip count',
    'date_span', 'integer','time span in days',
    'orig_adm0', 'character', 'name of highest administration level of origin location (Country)',
    'orig_adm1', 'character', 'name of administration level 1 of origin location (e.g. Division, State)',
    'orig_adm2', 'character', 'name of administration level 2 of origin location (e.g. District, County)',
    'orig_adm3', 'character', 'name of administration level 3 of origin location (e.g. Sub-district, Province)',
    'orig_adm4', 'character', 'name of administration level 4 of origin location (e.g. City, Municipality)',
    'orig_adm5', 'character', 'name of administration level 5 of origin location (e.g. Town, Village, Community, Ward)',
    'orig_type', 'character', 'administrative type for the origin location (e.g. sub-district, community vs town, or urban vs rural)',
    'orig_x', 'numeric', 'longitude of origin location centroid in decimal degrees (centroid of smallest admin unit',
    'orig_y', 'numeric', 'latitude of origin location centroid in decimal degrees (centroid of smallest admin unit)',
    'orig_pop', 'numeric', 'population size of lowest administrative unit for origin location',
    'dest_adm0', 'character', 'name of highest administration level of destination location (Country)',
    'dest_adm1', 'character', 'name of administration level 1 of destination location (e.g. Division, State)',
    'dest_adm2', 'character', 'name of administration level 2 of destination location (e.g. District, County)',
    'dest_adm3', 'character', 'name of administration level 3 of destination location (e.g. Sub-district, Province)',
    'dest_adm4', 'character', 'name of administration level 4 of destination location (e.g. City, Municipality)',
    'dest_adm5', 'character', 'name of administration level 5 of destination location (e.g. Town, Village, Community, Ward)',
    'dest_type', 'character', 'administrative type for the destination location (e.g. sub-district, community vs town, or urban vs rural)',
    'dest_x', 'numeric', 'longitude of destination location in decimal degrees (centroid of smallest admin unit)',
    'dest_y', 'numeric', 'latitude of destination location centroid in decimal degrees (centroid of smallest admin unit)',
    'dest_pop', 'numeric', 'population size of lowest administrative unit for destination location',
    'trips', 'numeric', 'total number of observed trips made from origin to destination during time span'
  ),
  ncol=3,
  byrow=TRUE)

colnames(tab) <- c('Variable', 'Class', 'Description')

knitr::kable(tab)

```

This data template can be populated by starting with the \code{travel_data_template} object and adding rows.

```{r trips}
# Travel among some locations
trip <- travel_data_template

n <- 30 # number of locations
trip[1:n,] <- NA # add rows for each location

# Time span of travel data
trip$date_start <- as.Date("2020-01-01")
trip$date_stop <- trip$date_start + 7
trip$date_span <- difftime(trip$date_stop, trip$date_start, units='days')

# Origin info: some counties within the same state
trip$orig_adm0 <- trip$dest_adm0 <- 'A' # Country
trip$orig_adm1 <- trip$dest_adm1 <- 'B' # State
trip$orig_adm2 <- sample(LETTERS, n, replace=T)
trip$dest_adm2 <- sample(LETTERS, n, replace=T)
trip$orig_type <- trip$dest_type <- 'County' # Type of admin unit for lowest admin level

# Some fake coordinates in decimal degrees
trip$orig_x <- rnorm(n, -90, 2)
trip$orig_y <- rnorm(n, 30, 1)
trip$dest_x <- rnorm(n, -90, 2)
trip$dest_y <- rnorm(n, 30, 1)

# Population sizes of the origins and destinations
trip$orig_pop <- rnbinom(n, size=5, mu=5000)
trip$dest_pop <- rnbinom(n, size=10, mu=10000)

trip$trips <- rnbinom(n, size=1, mu=100) # Number of reported trips
trip <- trip[!(trip$orig_adm2 == trip$dest_adm2),]
```



```{r stays}
# Stays in home location
stay <- travel_data_template
origins <- unique(c(trip$orig_adm2, trip$orig_adm2)) # all the
stay[1:length(origins),] <- NA

# Time span of travel survey
stay$date_start <- trip$date_start[1]
stay$date_stop <- trip$date_stop[1]
stay$date_span <- difftime(stay$date_stop, stay$date_start, units='days')

stay$orig_adm0 <- stay$dest_adm0 <- 'A' # Country
stay$orig_adm1 <- stay$dest_adm1 <- 'B' # State
stay$orig_adm2 <- stay$dest_adm2 <- origins
stay$orig_type <- stay$dest_type <- 'County'

for (i in 1:length(origins)) {

  sel <- which(trip$orig_adm2 == stay$orig_adm2[i])[1]
  stay$orig_x[i] <- stay$dest_x[i] <- trip$orig_x[sel]
  stay$orig_y[i] <- stay$dest_y[i] <- trip$orig_y[sel]
  stay$orig_pop[i] <- stay$dest_pop[i] <- trip$orig_pop[sel]
}

# Number of reported trip within home county
stay$trips <- rnbinom(length(origins), size=10, mu=1000)

# Combine
travel_data2 <- dplyr::full_join(trip, stay)

```
