---
title: "Fitting and simulating travel probability"
output: rmarkdown::html_vignette
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amssymb}
   - \usepackage{amsfonts}
vignette: >
  %\VignetteIndexEntry{Fitting and simulating travel probability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(mobility)
library(ggplot2)
library(reshape2)
library(viridis)
library(ggstance)
library(foreach)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  autodep=TRUE
)

set.seed(1234)
```

### Build data vectors
The travel probability model requires two pieces of information: the number of trips for the time interval that travel outside the origin location ($y_i$) and the total number of trips observed for that origin ($m_i$). Both vectors must be the same length but they can have missing values (`NA`) that represent an unobserved origin location. The `get_stay_data()` function can be used to build trips for each location that are aggregated to a desired spatial scale (e.g. admin level 1 or admin level 2).
```{r data}
stay_data <- get_stay_data(travel_data_sim, agg_adm=2)
y <- setNames(stay_data$travel, stay_data$orig_id)
m <- setNames(stay_data$total, stay_data$orig_id)

n_orig <- length(y)
miss <- sample(1:n_orig, n_orig*0.2) # missing observations
y[miss] <- m[miss] <- NA

y
m
```


### Estimate probability of travel for each location
To estimate the probability of traveling outside the origin location $i$, the `fit_prob_travel()` function uses a Beta-Binomial model with hierarchical structure to infer travel probability in unobserved locations.
$$
y_i \sim \text{Binom}(\tau_i, \sum_{\forall j} m_{ij})
$$
The random variable $y_i$ is the observed number of trips that leave origin $i$ within the time interval. Binomial parameters $\tau_i$ and $\sum_{\forall j} m_{ij}$ are the success probability and total number of observed trips emanating from origin $i$ respectively. 
$$
\begin{aligned}
\tau_i &\sim \text{Beta}(1+\alpha, 1+\beta) \\
\tau_\text{pop} &\sim \text{Beta}(1+\bar{\alpha}, 1+\bar{\beta})
\end{aligned}
$$
Binomial probabilities for each origin $\tau_i$ are drawn from a Beta distributed prior with shape and rate parameters $\alpha$ and $\beta$. The hierarchical structure comes from estimating $\alpha$ and $\beta$ as population-level hyper-priors for the origin-level probabilities $\tau_i$ and allowing $\tau_\text{pop}$ to inherit the overall population-level distribution denoted as $\bar{\alpha}$ and $\bar{\beta}$.
$$
\begin{aligned}
\alpha &\sim \text{Gamma}(0.01, 0.01) \\
\beta &\sim \text{Gamma}(0.01, 0.01)
\end{aligned}
$$  

This structure allows origin locations to have probabilities $\tau_i$ which are driven by data in each location and all unobserved locations regress to the population mean $\tau_\text{pop}$.
```{r fit1}
prob_trav <- fit_prob_travel(travel=y, total=m)
```
```{r, fig.width=5, fig.height=7}
prob_trav_smry <- summary(prob_trav, probs=c(0.025, 0.25, 0.75, 0.975))

ggplot(data=prob_trav_smry) +
  geom_point(aes(x=mean, y=names(y)), size=2) +
  ggstance::geom_linerangeh(aes(y=names(y), xmin=Q2.5, xmax=Q97.5)) +
  ggstance::geom_linerangeh(aes(y=names(y), xmin=Q25, xmax=Q75), size=1) +
  xlab('Probability of travel outside origin') + ylab('Origin') +
  xlim(0,1) +
  theme_bw() + theme(axis.text.x=element_text(size=10),
                     axis.text.y=element_text(size=10),
                     axis.title.x=element_text(size=12, margin = margin(t = 15)),
                     axis.title.y=element_text(size=12, margin = margin(r = 15)),
                     panel.border = element_rect(colour = "black", fill=NA, size=1),
                     legend.position='right')
```

### Simulate values of travel probability for each location
There can be relatively large confidence intervals around the travel probability estimated for locations without data. Since these locations inherit the population-level mean, uncertainty will likely be higher than locations with data. If we wanted to then use this model in a simulation, we may want to include this uncertainty in the model.

The `predict()` function can be used to simulate the fitted values of travel probability for each location. The function takes posterior estimates of the mean ($\mu$) and standard deviation ($\sigma$) for each origin $i$, and then derives the shape and rate parameters for the Beta distribution. Simulated values are random draws from this Beta distribution.

```{r}
# Simulate n realizations of travel probability for each location
predict(object=prob_trav, nsim=5, seed=123)
```

