---
title: "Fitting and simulating travel probability"
output: rmarkdown::html_vignette
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amssymb}
   - \usepackage{amsfonts}
vignette: >
  %\VignetteIndexEntry{Fitting and simulating travel probability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(mobility)
library(ggplot2)
library(reshape2)
library(viridis)
library(ggstance)
library(foreach)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  autodep=TRUE
)

set.seed(1234)
```

### Build data vectors
The travel probability model requires two pieces of information: the number of trips for the time interval that travel outside the origin location ($Y$) and the total number of trips observed for that origin ($N$). Both vectors must be the same length but they can have missing values (`NA`) that represent an unobserved origin location. The `get_stay_data` function can be used to build trips for each location that are aggregated to a desired spatial scale (e.g. admin level 1 or admin level 2).
```{r data}
stay_data <- get_stay_data(travel_data_sim, agg_adm=2)
Y <- setNames(stay_data$travel, stay_data$orig_id)
N <- setNames(stay_data$total, stay_data$orig_id)

n_orig <- length(Y)
miss <- sample(1:n_orig, n_orig*0.2) # missing observations
Y[miss] <- N[miss] <- NA

Y
N
```


### Estimate probability of travel for each location
To estimate the probability of traveling outside the origin location $i$, the `fit_prob_travel` function uses a Beta-Binomial model with hierarchical structure to infer travel probability in unobserved locations.
$$
y_i \sim \text{Binom}(\tau_i, n_i)
$$
The random variable $y_i$ is the observed number of trips that leave origin $i$ within the time interval. Binomial parameters $\tau_i$ and $n_i$ are the success probability and total number of observed trips in origin $i$ respectively. 
$$
\begin{aligned}
\tau_i &\sim \text{Beta}(\alpha, \beta) \\
\tau_\text{pop} &\sim \text{Beta}(\bar{\alpha}, \bar{\beta})
\end{aligned}
$$
Binomial probabilities for each origin $\tau_i$ are drawn from a Beta distributed prior with shape and rate parameters $\alpha$ and $\beta$. The hierarchical structure comes from estimating $\alpha$ and $\beta$ as population-level hyper-priors for the origin-level probabilities $\tau_i$ and allowing $\tau_\text{pop}$ to inherit the overall population-level distribution denoted as $\bar{\alpha}$ and $\bar{\beta}$.
$$
\begin{aligned}
\alpha &\sim \text{Gamma}(1,0.1) \\
\beta &\sim \text{Gamma}(1,0.1)
\end{aligned}
$$  

This structure allows origin locations to have probabilities $\tau_i$ which are driven by data in each location and all unobserved locations regress to the population mean $\tau_\text{pop}$.
```{r fit1}
prob_trav <- fit_prob_travel(travel=Y,
                             total=N,
                             DIC=TRUE,
                             format_locations=FALSE)

summarize_mobility(prob_trav)
```

To return estimates of travel probability for each location, the `format_locations` argument will summarize model parameters and return a data frame.
```{r fit2}
prob_trav <- fit_prob_travel(travel=Y,
                             total=N,
                             format_locations=TRUE)
head(prob_trav)
```
```{r, echo=FALSE, fig.width=5, fig.height=7}
ggplot(data=prob_trav) +
  geom_vline(aes(xintercept=Mean[ which(is.na(travel))[1] ]), color='red', linetype=2) +
  geom_point(aes(x=Mean, y=orig_id), size=2) +
  ggstance::geom_linerangeh(aes(y=orig_id, xmin=HPD2.5, xmax=HPD97.5)) +
  xlim(0,1) +
  xlab('Probability of travel outside origin') + ylab('Origin') +
  theme_bw() + theme(axis.text.x=element_text(size=9),
                     axis.text.y=element_text(size=9),
                     axis.title.x=element_text(size=12, margin = margin(t = 15)),
                     axis.title.y=element_text(size=12, margin = margin(r = 15)),
                     legend.position='right')
```

### Simulate values travel probability for each location
There can be relatively large confidence intervals around the travel probability estimated for locations without data. Since these locations inherit the population-level mean, uncertainty will likely be higher than locations with data. If we wanted to then use this model in a simulation, we may want to include this uncertainty in the model.

The `sim_prob_travel` function can be used to simulate values of travel probability for each location. Each time the function is called, it returns one random realization for each location.

```{r}
# Simulate one realization of travel probability for each location
sim_prob_travel(mu=prob_trav$Mean,
                sigma=prob_trav$SD,
                id=prob_trav$orig_id)

# Simulate n realizations of travel probability for each location
n <- 5
foreach(i=1:n, .combine=cbind) %do% {
  sim_prob_travel(mu=prob_trav$Mean,
                  sigma=prob_trav$SD,
                  id=prob_trav$orig_id)
}
```

